---
title: 'Topic Modeling in R: LDA vs NMF with Associated Press'
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

# Overview & Learning Goals

This R Markdown file presents a comparison of two topic modeling approaches using an R dataset:

- **Latent Dirichlet Allocation (LDA):** a probabilistic model that learns topics as word distributions and documents as mixtures of topics.
- **Non-negative Matrix Factorization (NMF): ** a linear-algebra approach that factorizes a document–term matrix into non-negative parts (“topics”) and document weights.

We will:

1. Use the **Associated Press (AP)** corpus bundled in the `topicmodels` package.
2. Build a **shared vocabulary** once and reuse it for both models.
3. Fit **LDA on counts** and **NMF (KL divergence) on V_counts** computed from the same counts.
4. Inspect **top words** per topic.
5. Compute two **model-agnostic** metrics:
   - **UMass coherence** (higher/less negative $\approx$ better internal consistency).
   - **Mean Jensen–Shannon distance (JSD)** between topics (higher = more distinct).

---

# Setup

## Packages

We use `topicmodels` (LDA + AP dataset), `NMF` (NMF), and base sparse-matrix tooling via `Matrix` and `slam`.

```{r setup, message=FALSE, warning=FALSE}
pkgs <- c("topicmodels", "NMF", "Matrix", "slam")
to_install <- setdiff(pkgs, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install, repos = "https://cloud.r-project.org")

# NMF depends on Biobase (from Bioconductor). Install if missing.
if (!requireNamespace("Biobase", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install("Biobase", ask = FALSE, update = FALSE)
}
if (!requireNamespace("tm", quietly = TRUE)) install.packages("tm")
library(tm)

library(topicmodels)
library(NMF)
library(Matrix)
library(slam)
set.seed(0)
```

## Configuration

```{r config}
n_topics     <- 10        # number of topics
n_top_words  <- 20        # words to display per topic
max_features <- 5000      # vocabulary cap for fairness and speed
min_df       <- 10        # drop very rare terms: min docs containing term
max_df_prop  <- 0.5       # drop very common terms: max doc proportion
random_state <- 0         # reproducibility
```

---

# Data

## Load the Associated Press corpus

`topicmodels` ships a preprocessed **DocumentTermMatrix** named `AssociatedPress` (AP).

```{r data}
data("AssociatedPress", package = "topicmodels")
dtm <- AssociatedPress  # class: DocumentTermMatrix (tm + slam)
n_docs <- nrow(dtm); n_terms <- ncol(dtm)
n_docs; n_terms
```

---

# Shared Vocabulary & Matrices (Fair Comparison)

We build **one** filtered vocabulary and use it for both models:

- **Counts (DTM)** → input to **LDA**
- **TF–IDF** (computed from those same counts) → input to **NMF**

Filtering steps:

- Remove **rare** terms (`min_df`) and **overly common** terms (`max_df_prop`).
- Cap to the `max_features` most frequent terms.
- Keep everything sparse.

```{r vocab, message=FALSE, warning=FALSE}
# Document frequency: number of documents containing the term
df <- slam::col_sums(dtm > 0)                                   # df_j = count of docs where term j appears at least once
keep <- (df >= min_df) & (df <= max_df_prop * n_docs)           # keep terms that are not too rare and not too common

# Limit to top features by overall term frequency (within 'keep')
tf  <- slam::col_sums(dtm[, keep])                              # total term frequency per kept term (sum over all docs)
ord <- order(tf, decreasing = TRUE)                             # rank kept terms by frequency (highest first)
if (length(ord) > max_features) ord <- ord[seq_len(max_features)] # cap to top 'max_features' indices if needed
keep_terms <- names(tf)[ord]                                    # get the term names (vocabulary) for those top features

# Apply selection
dtm_filt <- dtm[, keep_terms]                                   # restrict DTM to the selected terms (columns)
dtm_filt <- dtm_filt[, slam::col_sums(dtm_filt) > 0]            # drop any columns that ended up all-zero (safety)

# Drop empty documents
nz_docs <- slam::row_sums(dtm_filt) > 0                         # logical vector: which rows (docs) still have at least one term
dtm_filt <- dtm_filt[nz_docs, ]                                 # keep only non-empty documents

# Refresh doc count for downstream code
n_docs <- nrow(dtm_filt)                                        # update N to reflect the filtered corpus size

dim(dtm_filt)                                                   # show final matrix dimensions: (#docs, #terms)

```

## Build TF–IDF (sublinear TF + L2 row norm)

We approximate scikit-learn’s default TF–IDF:

- **TF:** sublinear scaling `log1p(count)`
- **IDF:** `log((1 + N) / (1 + df)) + 1`
- **Normalize:** L2 per document (row)

Sublinear scaling means replacing raw term frequency with a concave transform—typically **log(1 + tf)** (or sometimes $\sqrt{\text{tf}}$). It gives **diminishing returns** for repeated occurrences of the same word, so a term appearing 50 times isn’t treated 50× more important than one appearing 5 times. This **dampens bursty/repetitive words**, reduces bias toward longer documents, and generally yields more stable, interpretable weights in TF-IDF, retrieval, and topic modeling.

```{r tfidf}
# Convert DTM (slam::simple_triplet_matrix) to a sparse dgCMatrix (Matrix package)
stm <- dtm_filt

# Use existing document/term names if present; otherwise create synthetic IDs.
doc_ids  <- rownames(stm); if (is.null(doc_ids))  doc_ids  <- paste0("doc_",  seq_len(stm$nrow))
term_ids <- colnames(stm); if (is.null(term_ids)) term_ids <- paste0("term_", seq_len(stm$ncol))

# Build a column-compressed sparse matrix (dgCMatrix) from slam triplet slots:
#   stm$i -> row indices (docs)
#   stm$j -> column indices (terms)
#   stm$v -> counts (term frequencies)
# Provide dims and human-friendly dimnames for later inspection.
V_counts <- sparseMatrix(
  i = stm$i, j = stm$j, x = stm$v,
  dims = c(stm$nrow, stm$ncol),
  dimnames = list(Docs = doc_ids, Terms = term_ids)
)

# --- TF–IDF (sublinear TF + IDF + L2 row norm) ---

# Sublinear TF: replace raw counts tf with log(1 + tf).
# Access nonzero values directly via the @x slot for efficiency.
V_tf <- V_counts
V_tf@x <- log1p(V_tf@x)

# IDF on the filtered DTM:
# df_filt = #docs containing each term (document frequency).
df_filt <- slam::col_sums(dtm_filt > 0)

# Use the filtered doc count N. If you prefer this chunk self-contained,
# uncomment the next line and remove earlier n_docs definitions:
# n_docs <- nrow(stm)

# Smoothed IDF to avoid div-by-zero and overly extreme values:
# idf_j = log((1 + N) / (1 + df_j)) + 1
idf <- log((1 + n_docs) / (1 + as.numeric(df_filt))) + 1

# Apply IDF on the right: (docs x terms) %*% diag(idf) scales each term column by its IDF.
V_tfidf <- V_tf %*% Diagonal(x = idf)

# L2-normalize each document vector to unit length:
# row_norm_d = sqrt(sum_t (tfidf_{d,t}^2)).
row_norm <- sqrt(rowSums(V_tfidf^2))

# Guard against zero rows so scaling is safe.
row_norm[row_norm == 0] <- 1

# Left-multiply by diag(1/row_norm) to scale each row to unit L2 norm.
V_tfidf <- Diagonal(x = 1 / row_norm) %*% V_tfidf

# Report final matrix shape: (#docs, #terms)
dim(V_tfidf)
```

Note: A **DTM** is a **Document–Term Matrix**: a bag-of-words table where **rows are documents**, **columns are terms (tokens)**, and each cell holds a weight—usually a **count** (term frequency), but it could be binary or TF-IDF. It captures *what words appear how many times* in each document, while **ignoring word order** (bag-of-words).

* Shape: $N \times V$ (documents × vocabulary). Very sparse in practice, so we store it as a sparse matrix (e.g., `slam::simple_triplet_matrix`, `Matrix::dgCMatrix`).
* Conventions: **DTM** = docs as rows; **TDM** = the transpose (terms as rows).
* Modeling: **LDA** expects **integer counts** from a DTM; **NMF/LSA** can use counts or TF-IDF.

Here’s what each part is doing, in order:

1. **From slam triplets to a fast sparse matrix (counts)**

* `stm <- dtm_filt` takes your filtered **slam::simple_triplet_matrix** (docs × terms).
* `doc_ids`, `term_ids` ensure you have readable row/col names; if missing, synthetic IDs are created.
* `V_counts <- sparseMatrix(i=stm$i, j=stm$j, x=stm$v, ...)` rebuilds the matrix as a **Matrix::dgCMatrix** (column-compressed, efficient for linear algebra). The triplet slots mean: `stm$i` = document indices, `stm$j` = term indices, `stm$v` = count values.

2. **TF step: sublinear scaling of counts**

* `V_tf <- V_counts; V_tf@x <- log1p(V_tf@x)` applies **sublinear TF**: $tf'=\log(1+tf)$.
  This gives diminishing returns to repeated words and reduces document-length bias.

3. **IDF step: document frequency and smoothed IDF**

* `df_filt <- slam::col_sums(dtm_filt > 0)` computes document frequency for each term: $df_j=\#{d:,tf_{d,j}>0}$.
* `idf <- log((1 + n_docs) / (1 + df_filt)) + 1` computes smoothed IDF:
  $$idf_j=\log!\Big(\frac{1+N}{1+df_j}\Big)+1,$$
  where $N=n_{\text{docs}}$. The +1’s prevent division by zero and extreme values for rare terms.

4. **Build TF–IDF and L2-normalize each document**

* `V_tfidf <- V_tf %*% Diagonal(x = idf)` multiplies each **term column** by its $idf_j$, i.e., $tfidf_{d,j}=tf'_{d,j}\cdot idf_j$.
* `row_norm <- sqrt(rowSums(V_tfidf^2))` computes $\lVert \mathbf{x}_d\rVert_2$ for each document row.
* `row_norm[row_norm == 0] <- 1` safeguards empty rows.
* `V_tfidf <- Diagonal(x = 1 / row_norm) %*% V_tfidf` scales each document to unit length:
  $$\mathbf{x}_d \leftarrow \frac{\mathbf{x}_d}{\lVert \mathbf{x}_d\rVert_2}.$$

5. **Sanity check**

* `dim(V_tfidf)` prints the final shape (docs × terms), confirming TF–IDF has the same dimensions as the counts matrix.

---

# Fit the Models

## LDA on counts (variational EM)

**VEM (Variational EM) for LDA** is a deterministic inference method that alternates between a **variational E-step** (estimate per-document topic proportions and per-word topic assignments by maximizing the ELBO) and an **M-step** (update global topic–word distributions and sometimes hyperparameters). It’s fast, stable given a seed, and works directly with **count** DTMs.

```{r lda, message=FALSE}
control_lda <- list(estimate.alpha = TRUE, seed = random_state)
lda_model <- LDA(dtm_filt, k = n_topics, method = "VEM", control = control_lda)
lda_model
```

## NMF on counts (KL / ‘brunet’)

We use KL divergence via the Brunet multiplicative updates, initialized with NNDSVD, on the **count** matrix. This setting typically yields more interpretable topics on news corpora than Euclidean on TF–IDF, even with a single run.

The KL-divergence–based **multiplicative updates** algorithm (Brunet et al., 2004) that minimizes $D_{\mathrm{KL}}(V|WH)$. It’s well-suited to **count-like** data, tends to produce sparse, parts-based topics, and benefits from good initialization (e.g., **NNDSVD**) or a few replicates to avoid poor local minima.


```{r nmf, message=FALSE}
set.seed(random_state)
nmf_model <- suppressWarnings(
  nmf(as.matrix(V_counts), rank = n_topics, method = "brunet",
      nrun = 1, seed = "nndsvd")
)
nmf_model
```

---

# Topic Inspection

Extract **top words** for each topic.

```{r top-words, message=FALSE}
# LDA: topic-word probabilities (phi), dimensions: k x vocab
lda_post <- posterior(lda_model)
phi_lda  <- lda_post$terms

# NMF: basis matrix W (features x k); convert to topic-word (k x features) and row-normalize
W <- basis(nmf_model)
phi_nmf <- t(W)
phi_nmf <- phi_nmf / pmax(rowSums(phi_nmf), 1e-12)

feat_names <- colnames(as.matrix(dtm_filt))

top_words <- function(phi, topn = n_top_words, features = feat_names) {
  apply(phi, 1, function(row) {
    ids <- order(row, decreasing = TRUE)[seq_len(min(topn, length(row)))]
    paste(features[ids], collapse = " ")
  })
}

cat("### Topics in LDA (counts)\n")
tw_lda <- top_words(phi_lda, n_top_words, feat_names)
for (i in seq_along(tw_lda)) cat(sprintf("Topic #%d:\n%s\n\n", i-1, tw_lda[[i]]))

cat("### Topics in NMF (counts, KL)\n")
tw_nmf <- top_words(phi_nmf, n_top_words, feat_names)
for (i in seq_along(tw_nmf)) cat(sprintf("Topic #%d:\n%s\n\n", i-1, tw_nmf[[i]]))
```

1. **Get LDA topic–word probs ($\phi$)**

* `lda_post <- posterior(lda_model)` computes posterior estimates for your fitted LDA model.
* `phi_lda <- lda_post$terms` pulls the **topics × vocabulary** matrix where each row is a topic and each entry is $P(\text{word}\mid \text{topic})$. Rows already sum to 1 (probabilities).

2. **Get NMF topic–word weights and make them comparable**

* `W <- basis(nmf_model)` returns the NMF **basis** matrix $W$ with shape **features × k** (terms × topics).
* `phi_nmf <- t(W)` transposes to **topics × features** so rows correspond to topics like LDA’s $\phi$.
* `phi_nmf <- phi_nmf / pmax(rowSums(phi_nmf), 1e-12)` row-normalizes so each topic’s weights sum to $\approx 1$ (now interpretable like probabilities). `pmax(..., 1e-12)` avoids division by zero.

3. **Name the columns (vocabulary)**

* `feat_names <- colnames(as.matrix(dtm_filt))` grabs the term names in the same order as your matrices’ columns (the vocabulary you filtered earlier).

4. **Helper to extract top words per topic**

* `top_words <- function(phi, topn, features)`:

  * `apply(phi, 1, ...)` iterates **row-wise** (topic by topic).
  * `ids <- order(row, decreasing = TRUE)[seq_len(min(topn, length(row)))]` finds the indices of the **top-$N$** highest-weight words for that topic.
  * `paste(features[ids], collapse = " ")` maps those indices to term strings and joins them into one label line.
  * Returns a character vector: one string per topic.

5. **Print LDA topics**

* `cat("### Topics in LDA (counts)\n")` prints a header.
* `tw_lda <- top_words(phi_lda, ...)` gets the top words from LDA’s $\phi$.
* The `for` loop prints “Topic #0, #1, …” (note the `i-1` to show **0-based** labels) and the corresponding top words.

6. **Print NMF topics**

* `cat("### Topics in NMF (counts, KL)\n")` prints a header that matches your configuration.
* `tw_nmf <- top_words(phi_nmf, ...)` gets top words from the normalized NMF rows.
* The loop prints each topic index (again zero-based) and its top words.

**Notes:**

* LDA’s `phi_lda` is **already probabilistic**; for NMF we **normalize** so both methods’ topic rows can be read similarly.
* Topic indices are arbitrary within each model — **Topic #3 in LDA** isn’t guaranteed to match **Topic #3 in NMF**. If you ever compare across models/runs, align topics first.

---

# Model-Agnostic Metrics

## UMass Coherence

**Question:** Do a topic’s top words tend to co-occur in the same documents?  

We compute a log-based co-occurrence score across top-word pairs (less negative $\approx$ better).

```{r coherence, message=FALSE}
umass_coherence <- function(phi, dtm_counts, topn = 20, eps = 1e-12) {
  # Convert to sparse matrix (dgCMatrix)
  stm <- dtm_counts
  X <- sparseMatrix(i = stm$i, j = stm$j, x = stm$v,
                    dims = c(stm$nrow, stm$ncol))
  # Binarize
  X@x <- ifelse(X@x > 0, 1, 0)
  df <- Matrix::colSums(X)
  C  <- Matrix::t(X) %*% X

  # Top indices per topic
  top_ids <- apply(phi, 1, function(row)
    order(row, decreasing = TRUE)[seq_len(min(topn, length(row)))] )

  scores <- c()
  for (k in seq_len(nrow(phi))) {
    ids <- top_ids[, k]
    s <- 0; pairs <- 0
    if (length(ids) >= 2) {
      for (i in 2:length(ids)) {
        wi <- ids[i]
        for (j in 1:(i-1)) {
          wj <- ids[j]
          co <- C[wi, wj]
          s <- s + log((as.numeric(co) + eps) / (as.numeric(df[wj]) + eps))
          pairs <- pairs + 1
        }
      }
    }
    scores <- c(scores, ifelse(pairs > 0, s / pairs, NA_real_))
  }
  mean(scores, na.rm = TRUE)
}
```

## Mean Jensen–Shannon Distance (JSD) Between Topics

**Question:** How different are topics from each other overall?  

Treat each topic as a probability distribution over terms; average the Jensen–Shannon divergence across topic pairs (higher = more distinct).

```{r jsd, message=FALSE}
row_normalize <- function(M, eps = 1e-12) {
  M / pmax(rowSums(M), eps)
}

kl_div <- function(p, q, eps = 1e-12) {
  p <- p + eps; q <- q + eps
  sum(p * log(p / q))
}

js_div <- function(p, q, eps = 1e-12) {
  m <- 0.5 * (p + q)
  0.5 * kl_div(p, m, eps) + 0.5 * kl_div(q, m, eps)
}

mean_js_distance <- function(phi) {
  phi <- row_normalize(phi)
  K <- nrow(phi)
  d <- c()
  for (i in 1:(K-1)) {
    for (j in (i+1):K) {
      d <- c(d, js_div(phi[i, ], phi[j, ]))
    }
  }
  mean(d)
}
```

## Compute and Report Metrics

```{r metrics}
lda_coh <- umass_coherence(phi_lda, dtm_filt, topn = n_top_words)
nmf_coh <- umass_coherence(phi_nmf, dtm_filt, topn = n_top_words)

lda_js  <- mean_js_distance(phi_lda)
nmf_js  <- mean_js_distance(phi_nmf)

cat("=== Summary Metrics ===\n")
cat(sprintf("LDA  coherence (UMass): %.4f   | separation (JSD): %.4f\n", lda_coh, lda_js))
cat(sprintf("NMF  coherence (UMass): %.4f   | separation (JSD): %.4f\n", nmf_coh, nmf_js))
```

---

# Interpretation

- **UMass coherence**: Do a topic’s top words tend to appear together in documents? (higher/less negative $\approx$ better).  
- **Mean JSD**: How different are the topics from each other overall? (higher = more distinct).  
- **Typical outcome**: LDA often has slightly better coherence; NMF (with narrower topics) often has higher separation.  
- Maintain a **shared vocabulary** to keep comparisons fair.

**UMass coherence** (higher/less-negative is better):

- LDA: **-1.794** pretty reasonable on AP; LDA topics (markets, USSR, Gulf War, courts, etc.) are visibly coherent.
- NMF: **-8.5110** very low coherence; the top words read “spiky/odd” and don’t co-occur much in documents. That’s typical when NMF is run with **Euclidean loss on TF–IDF**: it emphasizes rare/idiosyncratic terms rather than co-occurring ones.

**JSD separation** (higher = topics more dissimilar as distributions):

- NMF: **0.554** > LDA: **0.379**. NMF topics can become very “disjoint” (few shared high-weight words), so they look **far apart** even if they’re not coherent. High separation is not automatically good; coherence matters more for interpretability.
