---
title: "LDA and STM in R Example"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: inline
---

# Shared setup

### How the whole thing “flows”

1. **One preprocessing pipeline** (chunk 3) feeds **both** LDA and STM so we’re comparing apples to apples.
2. LDA section: **fit → top words → top docs → quick labels** (chunks 5–7).
3. STM section: **fit → labels → top docs → contrasts** (chunks 8–11), then a **no-content** STM to show the four scoring lists (chunk 12).

## 1) `setup`

* **What it does:** Quiet R Markdown messages and set a fixed random seed.
* **Why:** Keeps the PDF tidy and makes results reproducible.
* **Outputs:** Nothing user-facing; global knitr settings + `set.seed(42)`.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
set.seed(42)
```

## 2) *(packages)*

* **What it does:** Ensures required packages are installed, then loads them quietly.
* **Why:** Guarantees the rest of the file can run without missing libraries.
* **Outputs:** Loaded packages in memory (quanteda, topicmodels, stm, dplyr, kableExtra, etc.).

```{r}
# Ensure required packages are installed, then load them
options(repos = c(CRAN = "https://cloud.r-project.org"))

req_pkgs <- c(
  "quanteda", "quanteda.textstats", "quanteda.textmodels",
  "topicmodels", "SnowballC", "kableExtra",
  "dplyr", "tidyr", "stringr", "purrr", "Matrix",
  "readr", "tibble", "knitr", "stringi", "stm"
)

to_install <- setdiff(req_pkgs, rownames(installed.packages()))
if (length(to_install)) {
  install.packages(to_install, dependencies = TRUE)
}

# Load quietly
invisible(lapply(req_pkgs, function(p) {
  suppressPackageStartupMessages(library(p, character.only = TRUE))
}))
```

## 3) `preprocess_unified`

* **Inputs:** A raw text file `article.txt` (one article) in the working folder.
* **What it does (big picture):**

  * Sets **global knobs**: number of topics `K`, how many top words/docs to show, a frequency cutoff for pruning rare words, and whether to stem words.
  * Defines **domain stopwords** (e.g., “like”, “people”) to remove unhelpful filler.
  * **Reads and splits** the article into **paragraphs** (the “documents”) and builds simple **metadata** per paragraph: its index (`doc_id`), position in the article, length, and a coarse **section** label (early/middle/late).
  * Runs STM’s **`textProcessor`** and **`prepDocuments`** to tokenize, clean, and **trim** the vocabulary by frequency.
  * Builds two matrix “bridges” from the same cleaned data:

    * a **quanteda dfm** for quanteda workflows. `quanteda`is used only as a container/adapter: we wrap the STM-prepped counts into a `dfm` (`dfm_from_prep`) and then `convert()` that to a `topicmodels::DocumentTermMatrix` (`dtm_topicmodels`).
    * a **topicmodels DTM** for `topicmodels::LDA`. `topicmodels` is used to **fit LDA** on `dtm_topicmodels` and to extract results (`posterior(lda_fit)$topics` for $\theta$, `terms(lda_fit, n)`/`posterior(...)$terms` for top words).
    
  * Saves a vector that maps **kept rows back to the original paragraph IDs**.
* **Outputs (named objects):**

  * `prep$documents`, `prep$vocab`, `prep$meta` (for STM).
  * `dfm_from_prep` (quanteda dfm) and `dtm_topicmodels` (for LDA).
  * `paragraphs` (the cleaned text vector) and `orig_ids` (row→original paragraph map).
* **Used next by:** LDA fit (chunk 5) and all STM chunks (8–12).

The part about 2 matrix "bridges" is a bit confusing. By “matrix bridge,” I mean a tiny adapter layer that takes one canonical bag-of-words count table and presents it in whatever container a library expects—so every library sees the same numbers and vocabulary even though they use different R classes.

In the pipeline, STM’s `prepDocuments()` doesn’t give a rectangular table; it gives per-document term indices and counts. It first reassembles those into a single sparse D×V (documents x vocab) document-term matrix (rows = kept paragraphs, columns = the shared prep$vocab). That sparse matrix is the canonical representation of the data. From there, we create bridges:

* wrap the same counts as a quanteda dfm (so quanteda tools work), and
* wrap the same counts as a topicmodels::DocumentTermMatrix (so topicmodels::LDA() works).

All three objects: the sparse matrix, the dfm, and the DTM hold identical counts with identical column order; they’re just different “jackets” around the same numbers. That’s the whole point of the bridge: no re-tokenizing, no re-pruning, no drift. LDA and STM compare apples to apples.

**Why save `orig_ids`?**

`prepDocuments(...)` **drops** any paragraphs that become empty after cleaning/trimming and reorders metadata to match the kept documents. That means “row 17” in `prep$documents` is not necessarily “paragraph 17” from the original `paragraphs` vector.

We set `orig_ids <- prep$meta$doc_id`. Now:

* `orig_ids[d]` tells us which **original paragraph index** ended up as **row d** in the kept matrix.
* Later, when we pick top-documents by topic using `theta_lda` or `stm_fit$theta` (both are D×K, aligned with `prep$documents`), we map back to human-readable text via `paragraphs[ orig_ids[doc_id] ]`. Without `orig_ids`, we’d print the wrong snippet after trimming.

### Coordinate list (also called triplet) representation

A tiny toy picture

Suppose `vocab = c("cat","dog","tree")` so `V=3`. Two kept paragraphs (`D=2`):

* doc 1 has “cat cat dog” -> term indices `j=[1,2]`, counts `x=[2,1]` (cat twice, dog once);
* doc 2 has “tree” -> `j=[3]`, `x=[1]`.

Flattening gives `i_idx = [1,1,2]`, `j_idx = [1,2,3]`, `x_val = [2,1,1]`.

To see how this works:

* **`i_idx = [1, 1, 2]`** — row (document) indices for each **non-zero** cell. Read it as: “there’s a non-zero in **doc 1**, another non-zero in **doc 1**, and a non-zero in **doc 2**.” (R is 1-based.) The reason doc 1 appears twice is that it has **two distinct terms** with non-zero counts.

* **`j_idx = [1, 2, 3]`** — column (term) indices into the vocabulary. Read alongside `i_idx`: the first non-zero is for **term 1 = cat** in doc 1; the second is **term 2 = dog** in doc 1; the third is **term 3 = tree** in doc 2.

* **`x_val = [2, 1, 1]`** — the counts for those (doc, term) cells: (doc1, cat) has count **2**, (doc1, dog) has **1**, (doc2, tree) has **1**.

Put them together (by position `k`) to see the triples:

* `k=1`: `(i=1, j=1, x=2)` → (doc1, cat) = 2
* `k=2`: `(i=1, j=2, x=1)` → (doc1, dog) = 1
* `k=3`: `(i=2, j=3, x=1)` → (doc2, tree) = 1

To build the sparse matrix, think of the three vectors as **triples** that tell us where to *add* counts in a blank D×V table (rows = documents, columns = vocab). Here D=2 (doc1, doc2) and V=3 (cat, dog, tree).

1. Take the first triple `(i=1, j=1, x=2)`.
   Put **2** in **row 1, col 1** → (doc1, cat) = 2.

2. Second triple `(i=1, j=2, x=1)`.
   Put **1** in **row 1, col 2** → (doc1, dog) = 1.

3. Third triple `(i=2, j=3, x=1)`.
   Put **1** in **row 2, col 3** → (doc2, tree) = 1.

All other cells stay **0** (because they never appear in the triplets). If a document–term pair appeared more than once in `(i_idx, j_idx)`, you’d **add** the `x_val`s together at that cell.

`sparseMatrix(i,j,x,dims=c(2,3))` yields

```
       cat dog tree
doc1     2   1    0
doc2     0   0    1
```
as.dfm(...) wraps that as a dfm (document feature matrix);  convert(..., to="topicmodels") wraps it as a DTM (document term matrix). DFM and DTM are basically synonyms and the difference is just how they're wrapped as implementations for the libraries. If the original paragraphs were length 3 and prep dropped paragraph 2, then orig_ids = c(1,3) so row 1 maps back to paragraphs[1], row 2 to paragraphs[3].

```{r preprocess_unified, echo=TRUE, message=FALSE, warning=FALSE}
# Global knobs with safe defaults
if (!exists("K"))            K <- 5                 # requested number of topics
if (!exists("N_TOP_WORDS"))  N_TOP_WORDS <- 10      # words to show per topic
if (!exists("N_TOP_DOCS"))   N_TOP_DOCS <- 2        # top docs/snippets per topic
if (!exists("LOWER_THRESH")) LOWER_THRESH <- 2      # drop tokens with total count < this
if (!exists("DO_STEM"))      DO_STEM <- FALSE       # don't apply stemming in tokenization by default

# Domain stopwords beyond the built-in English list that we found a lot in the article
if (!exists("domain_stop")) {
  domain_stop <- c(
    "like","people","human","humans","really","just",
    "make","made","many","much","things","thing",
    "today","officials","report","reported","says",
    "can","one"
  )
}

# Load or reuse paragraphs (one paragraph = one "document")
if (!exists("paragraphs")) {
  txt <- readr::read_file("article.txt")          
  paragraphs <- stringr::str_split(txt, "\\n\\s*\\n")[[1]] |> # split on blank lines
    stringr::str_replace_all("\\s+", " ") |>                  # normalize whitespace (ie make it all a space)
    stringr::str_trim()
  paragraphs <- paragraphs[paragraphs != ""]                  # drop empty paragraphs
}

# Build document-level metadata used by STM:
# - doc_id: stable ID of the original paragraph
# - position: order in article 
# - length: token count per paragraph
# - section: coarse terciles (early/middle/late) for content covariate
meta <- tibble::tibble(
  doc_id   = seq_along(paragraphs),
  position = seq_along(paragraphs),
  length   = stringr::str_count(paragraphs, "\\S+")
) |>
  dplyr::mutate(
    section = cut(
      position,
      breaks = stats::quantile(position, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("early", "middle", "late")
    )
  )

# Tokenize/clean with STM helpers; apply stemming + stopword removal + pruning later
proc <- stm::textProcessor(
  documents          = paragraphs,
  metadata           = meta,
  lowercase          = TRUE,
  removestopwords    = TRUE,
  removenumbers      = TRUE,
  removepunctuation  = TRUE,
  stem               = DO_STEM,
  customstopwords    = domain_stop
)

# Construct STM's internal sparse format and trim rare terms (lower.thresh)
# This may drop empty documents; it also reorders/filters to "kept" docs
prep <- stm::prepDocuments(proc$documents, proc$vocab, proc$meta,
                           lower.thresh = LOWER_THRESH)

# Map rows of kept docs back to original paragraph indices (for snippet lookup later)
orig_ids <- prep$meta$doc_id

# Bridges for LDA / quanteda
D <- length(prep$documents)
V <- length(prep$vocab)

# Each prep$documents[[d]] is a 2×m matrix: row1 = term indices, row2 = counts
nnz_per_doc <- vapply(prep$documents, ncol, integer(1))                     # nonzeros per doc
i_idx <- rep(seq_len(D), times = nnz_per_doc)                               # row indices (doc IDs)
j_idx <- unlist(lapply(prep$documents, function(m) m[1, ]), use.names = FALSE) # col indices (term IDs)
x_val <- unlist(lapply(prep$documents, function(m) m[2, ]), use.names = FALSE) # counts

# Canonical sparse document-term matrix with vocab as column names
# "Canonical" = the single, source-of-truth D×V count table: rows = kept documents, cols = vocabulary terms.
# It's "sparse" because only non-zero counts are stored; all other word–doc pairs are implicit zeros.
# Column names are exactly prep$vocab, fixing the column order so every downstream step sees identical features.
# This matrix is the basis; the quanteda dfm and topicmodels DTM are just wrappers around the same counts.
# Building it from STM's per-doc indices ensures LDA and STM use identical tokenization and pruning.
dtm_sparse <- Matrix::sparseMatrix(
  i = i_idx,
  j = j_idx,
  x = x_val,
  dims = c(D, V),
  dimnames = list(NULL, prep$vocab)
)
# Two thin "bridges" (adapters) so other packages see the exact same counts/vocab
dfm_from_prep   <- quanteda::as.dfm(dtm_sparse)                         # for quanteda workflows
dtm_topicmodels <- quanteda::convert(dfm_from_prep, to = "topicmodels") # for topicmodels::LDA

# Objects produced:
# - prep$documents, prep$vocab, prep$meta  -> use for stm()
# - dfm_from_prep (quanteda dfm)           -> use for quanteda textmodels
# - dtm_topicmodels (topicmodels DTM)      -> use for topicmodels::LDA()
# - orig_ids                               -> use to index back to original paragraphs
```

## 4) `helpers`

This section provides two small utilities used throughout the tutorial. `top_docs_for_topic()` takes a **document–topic matrix $\theta$** and, for a chosen topic *k*, returns the **top N paragraphs** with short, readable snippets. Here $\theta$ is a D×K matrix of document–topic weights where $\theta_{d,k} \approx P(\text{topic k | document d})$; for LDA this comes from `posterior(lda_fit)$topics`, and for STM it’s `stm_fit$theta`. `fmt_table()` centralizes table styling (captions, alignment, column widths) so every table looks consistent without repeating kable/kableExtra boilerplate.

* **What it does:** Defines two utilities:
  * `top_docs_for_topic()`: given a topic’s document-probabilities, it finds the **top N paragraphs** for that topic and makes **short snippets**.
  * `fmt_table()`: one place to style kable tables consistently (caption, widths, alignment).
* **Outputs:** Two helper functions in memory.
* **Used next by:** LDA/STM “top docs” tables and most tables that follow.

```{r helpers}
# Helper: return the top-N documents for topic k, with human-readable snippets.
# Inputs:
#   - theta_mat: D×K document–topic matrix (theta), e.g. posterior(lda_fit)$topics or stm_fit$theta
#   - k:         1-based topic index to rank by
#   - topn:      how many documents to return
#   - map_ids:   optional integer vector of original paragraph IDs (e.g., orig_ids);
#                if provided, results report those original IDs instead of row indices
#   - paras:     character vector of paragraph texts indexed by original IDs (for snippets)
# Output: tibble with original Doc ID, the topic probability theta_{d,k}, and a 260-char snippet
top_docs_for_topic <- function(theta_mat, k, topn = 2, map_ids = NULL, paras = paragraphs) {
  # Rank documents by theta_{d,k} descending and keep the top N (guard against short corpora)
  ord <- order(theta_mat[, k], decreasing = TRUE)
  ord <- ord[seq_len(min(topn, length(ord)))]

  # Choose which IDs to display: original paragraph IDs if provided, else row numbers
  ids <- if (is.null(map_ids)) ord else map_ids[ord]

  # Build short, readable snippets from the original paragraphs (indexed by original IDs)
  snips <- ifelse(nchar(paras[ids]) > 260,
                  paste0(substr(paras[ids], 1, 260), "…"),
                  paras[ids])

  tibble(
    `Doc ID` = ids,
    prob     = round(theta_mat[ord, k], 3),  # theta_{d,k} for transparency
    snippet  = snips
  )
}

# Column-width preset used by 4-column tables (e.g., topic, doc id, prob, snippet)
WIDTHS_4 <- c("1.5cm","1.5cm","1.5cm","10cm")

# Single source of truth for table formatting so all tables share a consistent style.
# - caption:     table caption
# - align:       optional vector of column alignments; default centers all but last (left)
# - widths:      optional vector of explicit column widths (same length as ncol(df))
# - latex_opts:  kableExtra options; "hold_position" helps LaTeX place longtables sensibly
fmt_table <- function(df, caption, align = NULL, widths = NULL,
                      latex_opts = c("striped","hold_position")) {
  # Default alignment: center all columns except the last, which is left (better for text)
  if (is.null(align)) {
    n <- ncol(df)
    align <- if (n >= 2) c(rep("c", n - 1), "l") else rep("c", n)
  }

  # Build the kable and apply styling once here (DRY)
  out <- kbl(df, caption = caption, booktabs = TRUE, escape = TRUE,
             longtable = TRUE, align = align) |>
         kable_styling(latex_options = latex_opts)

  # Optional fixed widths (defensive check so widths match columns)
  if (!is.null(widths)) {
    stopifnot(length(widths) == ncol(df))
    for (j in seq_along(widths)) out <- column_spec(out, j, width = widths[[j]])
  }
  out
}
```


## Purpose

We're going to dive into something very similar to problem 1 from the homework using **R** with an article (each paragraph = one document), run a single preprocessing pipeline, then fit **both** **LDA** (topicmodels) **and** **STM** (with prevalence and a content covariate), using the **same number of topics K** as in Python, and compare their outputs.

* For **LDA**: Top **10 words** per topic; **Top 2 paragraphs** most associated with each topic; a short **2–3 word label** per topic (auto-derived, editable).
* For **STM**: Top **10 words** per topic (SAGE/FREX as appropriate); **Top 2 paragraphs** per topic; content-level **perspectives** across early/middle/late; plus an appendix with **Prob / FREX / Lift / Score** lists.

Topics are ordered by **prevalence** so LDA and STM tables share a consistent topic numbering for comparison.

---

# LDA (topicmodels)

Set **K** as we previously used for this problem in Python. Fit a baseline LDA on the trimmed DTM and order topics by prevalence so all downstream tables use a consistent topic order.

## 5) `fit_lda`

This step fits a baseline LDA to the trimmed `dtm_topicmodels` using the same **K** as in the Python run. The Gibbs sampler is fixed-seed for reproducibility. After fitting, the **document–topic weight matrix** from `posterior(lda_fit)$topics` is used to compute a simple **prevalence order** (mean topic weight across documents) so every downstream table refers to “Topic 1, 2, …” consistently. Top words per topic are then extracted and re-ordered to match that prevalence.

* **Inputs:** `dtm_topicmodels` from preprocessing and `K`.
* **What it does:** Fits an **LDA** model (Gibbs sampling), then:

  * Gets **posterior** topic probabilities per document (`theta_lda`).
  * Orders topics by **prevalence** (most common first).
  * Extracts **top 10 words** per topic and reorders columns to match that prevalence order.
  * Prints a **“Top 10 words per topic”** table.
* **Outputs:** `lda_fit`, `theta_lda`, `k_order`, and a top-words table.
* **Used next by:** LDA “top documents” and auto-labels.

```{r fit_lda}
K <- 5  # keep K consistent with a previous the Python run

# Gibbs sampler controls:
# - seed: reproducibility
# - alpha: document–topic prior (smaller => sparser mixtures)
# - burnin/iter/thin: warmup, total iterations, and thinning rate
ctrl <- list(seed = 42, alpha = 0.5, estimate.beta = TRUE,
             verbose = 0, burnin = 2000, iter = 1000, thin = 100)

# Fit baseline LDA on the trimmed DTM (same counts/vocab that STM saw)
lda_fit <- LDA(dtm_topicmodels, k = K, method = "Gibbs", control = ctrl)

# Cache posterior once for the whole LDA section:
# - post_lda$topics: D×K document–topic weights (approx P(topic k | document d))
# - post_lda$terms : K×V topic–term probabilities (not used directly below)
post_lda  <- posterior(lda_fit)
theta_lda <- post_lda$topics

# Order topics by corpus prevalence so all later tables share the same numbering.
# (Prevalence = column mean of the document–topic weights.)
k_order <- order(colMeans(theta_lda), decreasing = TRUE)

# Extract top 10 words per topic and reorder columns to match the prevalence order
terms_mat_lda     <- terms(lda_fit, 10)                    # words × topics (topics are columns)
terms_mat_lda_ord <- terms_mat_lda[, k_order, drop = FALSE]

# Build a compact table: topic index (by prevalence) + space-joined top words
top_terms_lda <- tibble(
  topic     = seq_along(k_order),                          # 1..K in prevalence order
  top_words = apply(terms_mat_lda_ord, 2, function(col) paste(col, collapse = " "))
)

# Render the summary; downstream chunks (top docs, labels) use the same topic order
kable(top_terms_lda, caption = "Top 10 words per topic (prevalence order descending)")
```

Reading down the top-word lists: Topic 1 centers on computing/brain capability (computer, brain, intelligence, machine, power). Topic 2 mixes conversational tokens (’re/’ll) with life/species/ASI. Topic 3 is about AGI/ASI and expert debate (ASI/AGI, likely, question, Bostrom). Topic 4 is the Kurzweil-style time/progress theme (future, time, years, progress). Topic 5 anchors the Turry vignette and goal/misalignment (turry, goal, systems, ani). 

---

## Most-associated paragraphs (top 2 per topic)

We use posterior **P(topic | doc)** to rank paragraphs, then show short snippets. Pick representative paragraphs by ranking docs with P(topic | doc); show short snippets (map to original paragraph IDs).

## 6) `lda_top_docs`

* **Inputs:** `theta_lda`, `k_order`, `orig_ids`, `paragraphs`.
* **What it does:** For each (prevalence-ordered) topic, finds the **top 2 paragraphs** and shows **short snippets**.
* **Outputs:** Nicely formatted table of **top paragraphs per LDA topic**.
* **Used next by:** Human interpretation; no downstream dependencies.

```{r lda_top_docs}
# Reuse cached objects from fit-lda: theta_lda, k_order, and orig_ids/paragraphs
top_docs_lda <- purrr::map_dfr(seq_along(k_order), function(i) {
  k_orig <- k_order[i]
  top_docs_for_topic(theta_lda, k_orig, topn = 2, map_ids = orig_ids, paras = paragraphs) |>
    dplyr::mutate(topic = i) |>
    dplyr::select(topic, `Doc ID`, prob, snippet)
})

fmt_table(
  top_docs_lda,
  caption = "Top 2 paragraphs per topic (prevalence order)",
  align   = c("c","c","c","l"),
  widths  = WIDTHS_4
)
```

The top paragraphs make those themes concrete. Topic 1 pairs a passage on aging/time with a reflection on computers beating us at tasks like chess. Topic 2 juxtaposes the definition of existential risk with a “why people aren’t thinking about this” setup. Topic 3 surfaces the survey timing results for AGI/ASI. Topic 4 highlights Kurzweil’s accelerating-returns and the “Die Progress Unit” thought experiment. Topic 5 anchors on the Turry handwriting/practice scene.     

---

## Auto-labels (2–3 words)

This step creates quick, human-readable labels for each LDA topic so tables and figures can reference “Topic 1: …” with a short name. The idea is intentionally simple: take the first 2–3 tokens from each topic’s top-words list and join them with slashes. We can treat these as placeholders; they’re for navigation, not ground truth.

## 7) Labels

* **Inputs:** The LDA **top-words** table.
* **What it does:** Creates quick **topic labels** by taking the first 2–3 top words (e.g., “economy/market/price”).
* **Outputs:** A compact **auto-labels** table.
* **Used next by:** Human interpretation; no downstream dependencies.

```{r labels}
# Build short, human-readable labels from the LDA top-words table.
# Strategy: normalize spaces -> split into tokens -> take first up to 3 -> join with "/".
auto_labels <- top_terms_lda |>
  dplyr::mutate(
    # Collapse any multiple spaces to a single space (defensive cleanup)
    label_src = stringr::str_replace_all(top_words, "\\s+", " "),
    # Split into tokens and keep the first up to 3; join with "/" for compactness
    label = purrr::map_chr(strsplit(label_src, " "),
                               ~ paste(head(.x, 3), collapse = "/"))
  ) |>
  dplyr::transmute(topic, label)

# Render a compact index of topic labels.
kable(auto_labels, caption = "Auto labels (first 2–3 top words)")
```

The auto-generated labels match the themes we see elsewhere: Topic 1 (“computer/brain/intelligence”) is about machine capability vs. human cognition; Topic 2 (“’re/life/species”) blends conversational tokens (’re) with biology terms; Topic 3 (“asi/will/agi”) centers on AGI/ASI and “will”/timing language; Topic 4 (“future/time/kurzweil”) is the acceleration/timeline motif; Topic 5 (“turry/goal/way”) anchors the Turry vignette and goal-specification/misalignment thread.

---

### Notes

- Stemming vs. no stemming will change the top words (e.g., *intellig* vs *intelligence*). I've elected not to use stemming for readability and it didn't seem to improve results.
- If the vocabulary is **too small** or **too large**, adjust `lower.thresh` via `stm::prepDocuments(...)`
- That is, lower `LOWER_THRESH` above; e.g., `LOWER_THRESH <- 6` or `10`).  
- Results will **not** exactly match Python’s LDA due to different implementations/seeds, but they should be comparable.


# STM

The main three differences between `stm` and LDA are listed below. For our purposes, we only will focus on the second two.

* Priors. In LDA, document–topic proportions use a Dirichlet prior: $\theta_d \sim \mathrm{Dirichlet}(\alpha)$. In STM, they use a logistic–normal prior whose mean depends on metadata: $z_d \sim \mathcal N(X_d \gamma,\ \Sigma)$ and $\theta_d=\mathrm{softmax}(z_d)$. Here $X_d\gamma$ lets covariates shift topic prevalence, and $\Sigma$ induces correlations among topics.

* Prevalence covariates (affect how much of each topic shows up). Document-level predictors enter the mean of $z_d$ in the logistic–normal prior, so $\theta_d$ shifts with metadata. Use when you expect frequency changes across documents (time, author, party, length, etc.). R: `stm(..., prevalence = ~ party + s(year) + log(length))`.

* Content covariate (changes how a topic is worded). A single categorical variable creates group-specific word distributions $\beta_{k \mid g}$ for each topic $k$ and group $g$ (e.g., “immigration” may emphasize *asylum, rights* vs. *border, enforcement*). Use only when you truly expect vocabulary to differ by group. R: `stm(..., content = ~ party)`.


## 8) `Fit STM

Fit an STM on the same preprocessed corpus so results are comparable to LDA. The **prevalence** formula models where/when a paragraph appears and how long it is; the **content** formula allows each topic’s word usage to shift across sections of the article (early/middle/late). This yields topics that can change vocabulary by section while keeping a single topic space for the whole document set.

* **Inputs:** `prep$documents`, `prep$vocab`, `prep$meta`, and `K`.
* **What it does:** Fits STM with `prevalence = ~ scale(position) + scale(length)` and `content = ~ section` (factor with levels early/middle/late). Uses spectral initialization for stability.
* **Outputs:** `stm_fit` (the fitted model) and `K_stm` (topics actually learned; guards downstream code if some topics collapse).

```{r stm_fit, echo=TRUE}
# Fit STM on the same counts/vocab used by LDA
# - documents/vocab/meta: identical preprocessed inputs from prepDocuments(...)
# - K: requested number of topics (may be adjusted by the model if some die out)
# - prevalence: lets topic proportions vary with position in article and paragraph length
#   (scale() standardizes these predictors for numerical stability)
# - content: allows each topic's word usage to differ across the 'section' factor
#   (early/middle/late thirds of the article)
# - init.type="Spectral": robust initialization recommended by the stm authors
stm_fit <- stm(
  documents  = prep$documents,
  vocab      = prep$vocab,
  K          = K,
  prevalence = ~ scale(position) + scale(length),
  content    = ~ section,
  data       = prep$meta,
  init.type  = "Spectral"
)

# Record the number of topics actually present in the fit
# (useful if some requested topics collapse or are pruned)
K_stm <- ncol(stm_fit$theta)
```

### This is how STM uses covariates:

**What “prevalence” means here.**
The `prevalence` formula controls the topic proportions per document (how much of each topic shows up in a paragraph). In the metadata, `position` is the exact paragraph index in the article (1, 2, 3, …), and `length` is the token count of that paragraph. Writing `~ scale(position) + scale(length)` tells STM to use a logistic-normal regression where those two predictors explain why some paragraphs lean more toward certain topics than others. `scale()` simply centers and standardizes the predictors to help estimation (it doesn’t change their meaning).

**What “content” means here.**
The `content` formula controls the wording of topics, not how much of a topic a document has. `section` is a categorical factor we created by slicing `position` into terciles: “early”, “middle”, “late.” With `content = ~ section`, `section` is a content covariate. STM learns different word distributions for the same topic depending on the section. For example, a “risk” topic might emphasize general setup terms early, debate words in the middle, and consequence terms late.

**Why both `position` and `section`?**
They play different roles. `position` is a numeric, fine-grained signal that lets topic proportions drift smoothly across the article; `section` is a coarse, categorical signal that lets the **vocabulary** of each topic shift between early/middle/late. *content* (section) = changes **which** words a topic uses by location; *prevalence* (position/length) = changes in **how much** of each topic appears in a paragraph.

## 9) STM Labels (content-aware SAGE vs. FREX)

This chunk builds compact topic labels for the STM fit. If the model includes a content covariate (`section`), it uses `sageLabels()` and extracts *marginal* word lists (content-averaged), preferring **FREX** and falling back to **prob/lift/score** when needed. If there is no content covariate, it uses `labelTopics()`, which exposes **frex/prob/lift/score** at the top level. The returned matrices are normalized to “one row per topic,” then the top words are concatenated into a single string per topic. The caption records which path was used.

* **Inputs:** `stm_fit`, `N_TOP_WORDS` (and `K_stm`); derives a `has_content` flag from `stm_fit$beta$logbeta`.
* **What it does:** Detects whether the STM was fit **with a content covariate** and chooses the right label extractor.

  * **With content:** calls `sageLabels(stm_fit, n=...)` and pulls top words from `labs$marginal` (prefers **FREX**, falls back to **prob/lift/score**).
  * **Without content:** calls `labelTopics(stm_fit, n=...)` and pulls **frex** (falls back to **prob/lift/score**).
  * Normalizes shapes (transposes if needed), concatenates each topic’s top words into a single string, and builds `top_terms_stm = tibble(topic, top_words)`.
  * Renders a kable with a caption that reflects the path used: “SAGE marginal FREX” when content is present, “FREX” otherwise.
* **Outputs:** `top_terms_stm` tibble and a formatted **topic-labels table** that works for both STM setups.
* **Used next by:** Quick topic inspection/naming; no downstream dependencies.

```{r stm_labels, echo=TRUE}
# Detect whether the STM was fit WITH a content covariate.
# Multiple beta components imply multiple content levels.
has_content <- {
  lb <- stm_fit$beta$logbeta
  is.list(lb) && length(lb) > 1
}

# Get label object:
# - With content: SAGE labels (content-aware); returns $marginal and $cov.betas
# - Without:      labelTopics() (classic FREX/prob/lift/score at top level)
labs <- if (has_content) {
  sageLabels(stm_fit, n = N_TOP_WORDS)
} else {
  labelTopics(stm_fit, n = N_TOP_WORDS)
}

# Turn 'labs' into a K-row character vector of "w1 w2 ... wN" per topic.
# Handles both SAGE (labs$marginal$frex/prob/lift/score) and labelTopics
# (labs$frex/prob/lift/score). Transposes if topics are in columns.
get_words <- function(labs, K) {
  # SAGE path: prefer marginal FREX; otherwise fall back to prob/lift/score
  if (!is.null(labs$marginal)) {
    cand <- list(labs$marginal$frex, labs$marginal$prob,
                 labs$marginal$lift, labs$marginal$score)
    for (x in cand) if (!is.null(x)) {
      m <- as.matrix(x)
      if (nrow(m) != K && ncol(m) == K) m <- t(m)   # ensure rows = topics
      return(apply(m, 1, function(row) paste(row, collapse = " ")))
    }
    stop("SAGE labels found, but no marginal word matrices were present.")
  }

  # No-content path: labelTopics fields live at the top level
  cand <- list(labs$frex, labs$prob, labs$lift, labs$score)
  for (x in cand) if (!is.null(x)) {
    m <- as.matrix(x)
    if (nrow(m) != K && ncol(m) == K) m <- t(m)     # ensure rows = topics
    return(apply(m, 1, function(row) paste(row, collapse = " ")))
  }

  stop("No word lists found in labs object.")
}

# Build the table used in later displays (topic index + concatenated top words)
top_terms_stm <- tibble::tibble(
  topic     = seq_len(K_stm),
  top_words = get_words(labs, K_stm)
)

# Caption documents which extractor was used
caption <- if (has_content) {
  "STM: Top 10 words per topic (SAGE marginal FREX)"
} else {
  "STM: Top 10 words per topic (FREX)"
}

# Render a compact labels table
top_terms_stm |>
  kbl(caption = caption, booktabs = TRUE, longtable = TRUE, escape = TRUE,
      align = c("c","l")) |>
  kable_styling(latex_options = c("striped","scale_down")) |>
  column_spec(1, width = "1.2cm") |>
  column_spec(2, width = "12cm")
```

#### Results above

STM’s marginal FREX lists sharpen the same broad story. Topic 1 foregrounds brain/humanity and general setup terms. Topic 2 emphasizes computers/machines with Kurzweil and assessment words like median/likely. Topic 3 is “questions/time/power.” Topic 4 is the ASI/Bostrom/species/future line. Topic 5 pulls ANI/goal/nanobots/process.

#### SAGE

SAGE stands for Sparse Additive Generative (models of text), from Eisenstein, Ahmed, and Xing (2011). SAGE models each topic’s word usage as a background word distribution plus a sparse set of additive deviations. The words with the largest positive deviations are the most characteristic of that topic. In **stm**, `sageLabels()` uses this idea to produce topic word lists: it highlights words that are not just frequent, but distinctive relative to the background, and (when we have a content covariate) it can report both marginal lists and content-specific lists. Compared with raw probability lists, SAGE-style labels usually read more like true “signatures” of a topic.

#### What are “marginal” word lists?

In an STM with a content covariate (e.g., `section` = early/middle/late), the model learns a separate word distribution for each topic within each content level. A **marginal word list** is the single, overall word list for a topic obtained by **marginalizing over the content variable**—i.e., averaging the content-specific word weights across levels (typically weighted by how common those levels are in the corpus). It’s called *marginal* because it comes from the marginal (content-averaged) topic–word distribution, not the distributions conditioned on a specific content level. Intuitively: estimated with the covariate, summarized as if we didn’t condition on it.

#### Why “prefer FREX” and fall back to prob/lift/score?

Different rankings highlight different kinds of words:

* **FREX** balances **frequency** (words that actually appear often in the topic) and **exclusivity** (words that are distinctive to the topic). This usually yields the most **label-friendly** set—specific but not obscure, so it’s preferred.
* **prob** ranks by raw topic probability. It’s clear but can surface generic terms that also occur in other topics.
* **lift** favors words that are rare overall but common in the topic, so it highlights distinctive terms; sometimes these can be a bit niche.
* **score** is another distinctiveness-leaning ranking that often sits between prob and lift.

The code tries marginal FREX first. If that field isn’t available in the object returned by `sageLabels()`/`labelTopics()`, it falls back to marginal **prob**, then **lift**, then **score**, so we always get a sensible list even if some fields are missing.

## 10) STM: Top 2 Paragraphs per Topic

This mirrors the LDA “top documents” step but uses STM’s **document–topic weights** to rank paragraphs. Topics are first ordered by prevalence (mean weight across documents), then the top N paragraphs for each topic are pulled with `top_docs_for_topic()`. `orig_ids` maps rows in the STM fit back to the original paragraph indices so snippets display the correct text. The result is a compact table for quick inspection and side-by-side comparison with LDA.

* **Inputs:** `stm_fit$theta`, `orig_ids`, `paragraphs`.
* **What it does:** Same idea as LDA top docs, but using STM’s topic–document probabilities:

  * Orders topics by **prevalence** under STM.
  * Shows the **top 2 paragraphs** for each STM topic (with snippets).
* **Outputs:** A **“Top paragraphs per STM topic”** table.
* **Used next by:** Interpretation.

```{r stm_top_docs, echo=TRUE}
# Document–topic weights from the STM fit
theta_stm   <- stm_fit$theta

# Order topics by prevalence (mean document–topic weight), most prevalent first
k_order_stm <- order(colMeans(theta_stm), decreasing = TRUE)

# For each topic (in prevalence order), collect the top-N paragraphs and add a rank-based topic index
top_docs_stm <- purrr::map_dfr(seq_along(k_order_stm), function(i) {
  k_orig <- k_order_stm[i]  # original topic index in the STM fit
  top_docs_for_topic(
    theta_mat = theta_stm,
    k         = k_orig,
    topn      = N_TOP_DOCS,
    map_ids   = orig_ids,      # map back to original paragraph IDs for correct snippets
    paras     = paragraphs
  ) |>
    dplyr::mutate(topic = i, .before = 1)  # label topics by prevalence rank: 1..K
})

# Render a consistent, readable table (topic, Doc ID, prob, snippet)
fmt_table(
  top_docs_stm,
  caption    = sprintf("STM: Top %d paragraphs per topic (prevalence order)", N_TOP_DOCS),
  align      = c("c","c","c","l"),
  widths     = WIDTHS_4,
  latex_opts = c("striped","hold_position","scale_down")
)
```

The STM-ranked anchor paragraphs mirror the LDA themes but in STM’s prevalence order. Topic 1 links “stepping-stone/instrumental” goals with concrete ANI systems in cars. Topic 2 contrasts human bias about growth rates with the Müller-Bostrom timing survey. Topic 3 brings back Kurzweil’s acceleration argument alongside AGI median-year estimates. Topic 4 splits between Turry’s improving handwriting and a time-travel thought experiment. Topic 5 ties brain-level feasibility arguments to goal-specification/morality.     

## 11) STM: Wording by Content Level (Perspectives)

This section contrasts per-level wordings of each STM topic. Because the model was fit with a content covariate (section = early/middle/late), STM stores a separate topic-by-vocabulary matrix for each level. Sorting each matrix’s rows (topics) by their largest log word-weights yields top words per topic for each level, showing how the same topic shifts its wording between early, middle, and late sections.

* **Inputs:** `stm_fit` (needs a content covariate with levels, e.g., early/middle/late).
* **What it does:** Shows **contrasting word lists** across content levels for each topic (e.g., words that are more characteristic in “early” vs “late” sections).
* **Why:** Helps see how content covariates change topic wording.
* **Outputs:** A **perspectives/contrasts** table (topic × contrast with exemplar words).
* **Used next by:** Interpretation.

```{r stm_perspectives, echo=TRUE}
# Vocabulary and factor levels for the content covariate
vocab     <- stm_fit$vocab
lvl_names <- levels(prep$meta$section)

# For content models, stm_fit$beta$logbeta is a list:
# one K×V matrix per content level (rows = topics, cols = vocab terms).
logbetas  <- stm_fit$beta$logbeta

# Given a K×V log-weight matrix for a level, return the top-n words per topic,
# concatenated as a single string per topic (one string per row).
get_top <- function(logbeta, n = N_TOP_WORDS) {
  apply(logbeta, 1, function(row) {
    # order indices of vocab by decreasing weight, take top-n, map to tokens, join with spaces
    paste(vocab[order(row, decreasing = TRUE)][seq_len(min(n, length(row)))],
          collapse = " ")
  })
}

# Compute a list of character vectors: one vector per level, length = K topics (native STM order)
by_level <- lapply(logbetas, get_top, n = N_TOP_WORDS)

# Assemble a table with one column per content level (early/middle/late), native topic order
persp <- tibble::tibble(topic = 1:K_stm)
for (i in seq_along(lvl_names)) {
  persp[[ lvl_names[i] ]] <- by_level[[i]]
}

# Reorder topics by prevalence so this table matches other STM/LDA tables
# (prevalence = mean document–topic weight across documents)
k_order_stm <- order(colMeans(stm_fit$theta), decreasing = TRUE)
persp <- persp |>
  dplyr::slice(k_order_stm) |>
  dplyr::mutate(topic = dplyr::row_number(), .before = 1)  # renumber 1..K in prevalence order

# Render a compact contrasts table: each row = a topic; columns = level-specific top words
persp |>
  kbl(caption = "STM: Top 10 words per topic by content level (prevalence order)",
      booktabs = TRUE, longtable = TRUE, escape = TRUE,
      align = c("c", rep("l", length(lvl_names)))) |>
  kable_styling(latex_options = c("striped","scale_down"), font_size = 9) |>
  column_spec(1, width = "1.2cm") |>
  column_spec(2:(length(lvl_names) + 1), width = "3.2cm")
```

By section, wording shifts as expected. Early columns emphasize setup like “intelligence/able/build” and “time/evolution.” Middle columns lean into evaluation—Bostrom/ASI and survey markers such as “median,” along with “right/hard.” Late columns move to mechanisms and consequences, surfacing “nanobots,” “robotica,” and “handwriting,” plus goal/program language.     

## 12)  Four Labeling Criteria (No Content Fit)

This section refits STM **without** a content covariate to get a single, clean set of per-topic word rankings. It then contrasts the four common label criteria—**Prob**, **FREX**, **Lift**, **Score**—side by side. Using a no-content fit avoids per-level wording effects and makes the differences between ranking schemes easier to see. Topics are ordered by **prevalence** (mean document–topic weight) for consistent numbering.

```{r stm_fourlists, echo=TRUE}
# Refit STM WITHOUT a content covariate so we get a single set of lists
# (Prob / FREX / Lift / Score) per topic, unconfounded by per-level wording.
invisible(capture.output({
  stm_nocontent <- stm(
    documents  = prep$documents,
    vocab      = prep$vocab,
    K          = K,
    prevalence = ~ scale(position) + scale(length),  # allow topic proportions to vary by position/length
    data       = prep$meta,
    init.type  = "Spectral",
    seed       = 99
  )
}, type = "output"))  # silence sampler output in the knit

# Order topics by prevalence (mean document–topic weight) for consistency with other tables
K_nc       <- ncol(stm_nocontent$theta)
k_order_nc <- order(colMeans(stm_nocontent$theta), decreasing = TRUE)

# Pull the four ranking lists (each highlights a different notion of "top words")
labs4 <- labelTopics(stm_nocontent, n = N_TOP_WORDS)
# - labs4$prob : highest probability words within each topic
# - labs4$frex : words balancing frequency and exclusivity (label-friendly)
# - labs4$lift : words unusually concentrated in the topic vs. overall corpus
# - labs4$score: another distinctiveness-leaning ranking

# Helper: convert a labelTopics field into per-topic strings, respecting prevalence order.
# Handles both matrix (topics in rows OR columns) and list forms defensively.
col_strings <- function(x, k_order) {
  if (is.null(x)) return(rep("", length(k_order)))
  if (is.list(x)) {
    # List of length K: each element is a character vector of top words for a topic
    return(vapply(x[k_order], function(v) paste(v, collapse = " "), ""))
  }
  m <- as.matrix(x)
  if (ncol(m) == length(k_order)) {
    # Topics are columns: reorder columns by k_order, paste down each column
    apply(m[, k_order, drop = FALSE], 2, function(col) paste(col, collapse = " "))
  } else if (nrow(m) == length(k_order)) {
    # Topics are rows: reorder rows by k_order, paste across each row
    apply(m[k_order, , drop = FALSE], 1, function(row) paste(row, collapse = " "))
  } else {
    # Fallback: dimensions unexpected—return in given orientation
    apply(m, 2, function(col) paste(col, collapse = " "))
  }
}

# Build a compact table (prevalence-ordered topic index + the four label criteria)
terms_4 <- tibble::tibble(
  topic = seq_along(k_order_nc),                         # 1..K in prevalence order
  prob  = col_strings(labs4$prob,  k_order_nc),
  frex  = col_strings(labs4$frex,  k_order_nc),
  lift  = col_strings(labs4$lift,  k_order_nc),
  score = col_strings(labs4$score, k_order_nc)
)

# Render a side-by-side comparison of the four criteria (prevalence order)
terms_4 |>
  kbl(caption = "STM (no content): Top 10 words — Prob / FREX / Lift / Score (prevalence order)",
      booktabs = TRUE, longtable = TRUE, escape = TRUE,
      align = c("c","l","l","l","l")) |>
  kable_styling(latex_options = c("striped","scale_down"), font_size = 9) |>
  column_spec(1, width = "1.2cm") |>
  column_spec(2:5, width = "3.2cm")
```

This no-content STM run shows how the four ranking schemes surface different facets of the same topics. 

* **Prob** lists the highest-mass, on-theme words first. It's great for the gist but often a bit generic (e.g., “intelligence,” “machine,” “time,” “year”).

* **FREX** balances frequency with exclusivity, so it tends to elevate crisp, label-friendly cues that aren’t drowned across topics (e.g., survey/timing terms or proper names when present). 
* **Lift** amplifies tokens that are disproportionately concentrated in a topic relative to the whole corpus; that’s where rare but signature words pop up (think niche jargon, scenario terms, or named entities).
* **Score** lands between FREX and Prob in feel: more distinctive than raw probability but less quirky than Lift, often keeping summary-sounding verbs/nouns that read naturally in prose. Read across each topic row: the overlap between Prob and FREX gives the core label, while Lift/Score add the tell-tale “badges” that help you distinguish near-neighbors.

In short, use **FREX** for the label you print, sanity-check it against **Prob** for coverage, and mine **Lift/Score** for one or two distinctive add-ons when topics feel too similar.

* If a word is **high in Prob and FREX**, it’s part of the core theme—good label material.
* If it appears **only in Prob**, it’s likely generic glue (keep for interpretation, rarely for the printed label).
* If it appears **only in Lift**, it’s a sharp discriminator—great as a parenthetical qualifier (“AGI timelines (survey/median)”, “Goal-specification (tripwire)”), but avoid using Lift-only terms as the whole label.
* If **Score** agrees with FREX, that reinforces distinctiveness; if Score leans back toward Prob, the topic’s vocabulary is more shared, so rely on FREX + top paragraphs to avoid vague labels.
* When two topics look alike under Prob, check **Lift** for rare signals (proper names, scenario words) and **FREX** for balanced exclusives; those usually separate them immediately.
